---
title: Transfom a matrix of species presence absence into a PresenceAbsence object
author: Bruno Vilela
date: '2018-01-24'
slug: transfom-a-matrix-of-species-presence-absence-into-a-presenceabsence-object
categories:
  - macroecology
tags:
  - Grid
  - IUCN
  - letsR
  - Presence Absence Matrix
  - R
  - Shapefiles
---


Most of `letsR` package functions require a `PresenceAbsence` object [(see previous post)](https://rmacroecology.netlify.com/2018/01/23/a-guide-to-transform-species-shapefiles-into-a-presence-absence-matrix-based-on-a-user-defined-grid-system/). But some users may already have a matrix of species presence absence (PAM) generated somewhere else or created using the `lets.presab.grid` function. In this post I will show how to transform these PAM files into a `PresenceAbsence` object.

First, note that this will only be possible if your grid used to generate the PAM is a square or a rectangular grid. Most of `letsR` functions are based on raster files, and rasters can only be square or rectangular.

So, first I will generate an example file using the function `lets.presab.grid`.

```{r, message=FALSE, warning=FALSE}
library(letsR)
```

```{r}
sp.r <- rasterToPolygons(raster(xmn = -93, xmx = -29,
                                ymn = -57, ymx = 15,
                                resolution = 5))
# Give an ID to the cell
slot(sp.r, "data") <- cbind("ID" = 1:length(sp.r),
                            slot(sp.r, "data"))
data(Phyllomedusa)
projection(Phyllomedusa) <- projection(sp.r)
resu <- lets.presab.grid(Phyllomedusa, sp.r, "ID")
rich_plus1 <- rowSums(resu$PAM) + 1
colfunc <- colorRampPalette(c("#fff5f0", "#fb6a4a", "#67000d"))
colors <- c("white", colfunc(max(rich_plus1)))
plot(resu$grid, border = "gray40",
     col = colors[rich_plus1])
map(add = TRUE)
```

Now we have a PAM in `matrix` format, with the row names indicating the cell ID in a gird.
```{r}
# Print only te first rows and columns of the PAM
resu$PAM[1:5, 1:3]
```

And we also have a grid.
```{r}
resu$grid
```

We have to first transform the grid into a raster file.
To do this, you have to know the resolution and extent of your grid. In this case the resolution is 5, and we can get the extent direct from the grid. 
```{r}
ext <- extent(resu$grid)
xmn <- ext[1]
xmx <- ext[2]
ymn <- ext[3]
ymx <- ext[4]

r <- raster(resolution = 5, xmn = xmn, xmx = xmx,
            ymn = ymn, ymx = ymx)
rich <- rasterize(resu$grid, r, field = "ID")
values(rich) <- rowSums(resu$PAM)
plot(rich)
map(add = TRUE)
```

Now the grid is in a raster format and contains species richness values, which is one of the objects of the `PresenceAbsence` object. We can now make the other two.
```{r}
xy <- xyFromCell(rich, 1:ncell(rich))
colnames(xy) <- c("Longitude(x)", "Latitude(y)")
pam <- cbind(xy, resu$PAM)
pam[1:5, 1:3] # print first 3 columns and 5 rows
species_names <- colnames(resu$PAM)
```

We also have the PAM with the centroid of the cells and the species name file.
We are ready to make the `PresenceAbsence` object.
```{r}
PAM_from_grid <- list(pam, rich, species_names)
data("PAM")
names(PAM_from_grid) <- names(PAM)
class(PAM_from_grid) <- class(PAM)
```

The transformation is complete. Now we can use all the functions for a `PresenceAbsence` object.
```{r}
summary(PAM_from_grid)
plot(PAM_from_grid)
```



